The error in Bayesian phylogenetic reconstruction when speciation co-occurs

### 2020-02-06 article.tex
Do not use '\input' in file that is already added by an '\input' as it does not scale (that is, this TeX cannot be used as part of a thesis)
Do not '\input' in a file added by '\input', as it scales badly (that is, it renders the TeX unusable for a thesis)
'speciation completion' are two words
Add hypothesis section
Renamed section to 'hypotheses'
Changes Giappo

% Comments
\newcommand*\rampal[1]{\textcolor{red}{\textbf{[RSE: #1]}}}
\newcommand*\richel[1]{\textcolor{orange}{\textbf{[RJCB: #1]}}}
\newcommand*\gio[1]{\textcolor{blue}{\textbf{[GL: #1]}}}

@@

\hyphenation{
  BEAST
  Pa-ra-me-ter
  Drum-mond 
  Bayes-ian 
  Mr-Bayes 
  ap-proach-es 
  Rev-Bayes 
  cre-ate
  spe-ci-a-tion-com-ple-tion
  pro-trac-ted
 }

@@

\input{razzo-introduction.tex}

\input{razzo-methods.tex}

@@

>>>>

% Comments
\newcommand*\rampal[1]{\textcolor{red}{\textbf{[RSE: #1]}}}
\newcommand*\richel[1]{\textcolor{orange}{\textbf{[RJCB: #1]}}}
\newcommand*\giovanni[1]{\textcolor{blue}{\textbf{[GL: #1]}}}

@@

\hyphenation{
  BEAST
  Pa-ra-me-ter
  Drum-mond 
  Bayes-ian 
  Mr-Bayes 
  ap-proach-es 
  Rev-Bayes 
  cre-ate
  spe-ci-a-tion
  com-ple-tion
  pro-trac-ted
 }

@@

\input{razzo-introduction.tex}
\iffalse % probably not present in the final version
\input{razzo-hypotheses.tex}
\fi
\input{razzo-methods.tex}

@@

\input{fig_results.tex}

\input{razzo-discussion.tex}

### 2020-02-06 razzo-abstract.tex
Changes Giappo

% One or two sentences providing a basic
% introduction to the field,
% comprehensible to a scientist in any discipline.
There exist millions of species on Earth,
all originating from a common ancestor billions
of years ago.
The field of phylogenetics uses heritable material (e.g. DNA)
to determine the evolutionary history of species.

% Two to three sentences of
% more detailed background, comprehensible to
% scientists in related disciplines.
Starting from heritable material and explicit assumptions,
Bayesian phylogenetics allows to infer a jointly-estimated 
phylogeny and parameter estimates distribution.
One of these assumptions is the speciation model, which
mathematically describes the branching process of a phylogeny in time.
Speciation models commonly assume that speciation events are independent and disjointed. Yet, such assumption may overlook the complexity of certain processes, where environmental changes promote speciation events in multiple lineages.
% Purpose of the paper
This new layer of complexity can be captured developing a novel ad hoc speciation model. However, the introduction of a new model could be not necessary if current models are already capable of describing the process to a satisfactory extent, under simpler assumptions.
Here we investigate the extent of the discrepancy produced by current BEAST2's tree priors, when trying to infer back trees where speciation can co-occur.

@@

% Two or three sentences explaining what
% the main result reveals in direct
% comparison to what was thought to be the case
% previously, or how the main result adds to
% previous knowledge.
We show that the extent of the inference error grows with the amount of co-occuring speciation events to establish the limits of BEAST2's standard tree priors.

% One or two sentences to put the results into a
% more general context.
These results allow phylogeneticists to judge under which circumstances the commonly used speciation model can be safely used.

>>>>

% One or two sentences providing a basic
% introduction to the field,
% comprehensible to a scientist in any discipline.
Millions of species exist on Earth, all originating from a common ancestor billions
of years ago.
The field of phylogenetics uses heritable material (e.g. DNA) to determine the evolutionary history of species.

% Two to three sentences of
% more detailed background, comprehensible to
% scientists in related disciplines.
Starting from heritable material and explicit assumptions, Bayesian phylogenetics allows to infer a jointly-estimated phylogeny and parameter estimates distribution.
One of these assumptions is the speciation model, which mathematically describes how species are added or removed from a phylogeny in time.
Speciation models commonly assume that speciation events are independent and disjointed. Yet, such assumption may overlook the complexity of certain processes, where environmental changes promote speciation events in multiple lineages.
% Purpose of the paper
This new layer of complexity can be captured developing a novel ad hoc speciation model. However, the introduction of a new model could be not necessary if current models are already capable of describing the process to a satisfactory extent, under simpler assumptions.
Here we investigate the extent of the discrepancy produced by current BEAST2's tree priors, when trying to infer back trees where speciation can co-occur.

@@

% Two or three sentences explaining what
% the main result reveals in direct
% comparison to what was thought to be the case
% previously, or how the main result adds to
% previous knowledge.
We show that the extent of the inference error increases with the amount of co-occuring speciation events to establish the limits of BEAST2's standard tree priors in this circumstance.

% One or two sentences to put the results into a
% more general context.
These results allow phylogeneticists to judge under which circumstances the commonly used speciation model can be safely used. \giovanni{Not entirely sure if this is the best sentence to put here.}

### 2020-02-06 razzo-discussion.tex
Changes Giappo

We \richel{(at least I. If you do not expect this, let me know the 
reasoning why)} expected to find a bigger inference error 
when increasing the amount of co-occurring speciation events.
The reason is simple: the tree prior used assumes speciation does not
co-occur. The stronger the amount of co-occurring speciation, the stronger
the deviation from the species tree prior.

\subsection{Research}

@@

\subsection{Methodological choices}

When simulating a DNA alignment, we chose to use the simplest (JC69) site
model and the simplest (strict) clock model. One could argue these models
are overly simplistic and biologically irrelevant. We argue that this
actually improves the clarity of our results for two reasons. The first
reason is a practical one: because the computational load was lowered,
we were able to perform more replicates. For a process with this high
amount of stochaticity, we think this is highly favored. The second reason
is that we think this reduces the noise of our results, without 
hurting the experiment: in our setup, it is essential
to assume the same generative site and clock model as the one that is actually 
used. Sure, we could have generated the alignment with a 
complex (GTR) site model and a complex (relaxed log-normal) clock model, then
correctly assume these were the generative site and clock model.
But that would be besides the point of the experiment: to measure the impact
of the tree prior.

When simulating our phylogenies, we used an MBD model with and without
extinction. In the case in which extinction is present,
picking the BD tree prior as the generative tree prior seems justified:
it is the best-fitting standard tree prior. For the MBD model without
extinction, \richel{@Giappo: I think this is an interesting insight:} 
we could have picked the Yule tree prior. We \richel{at least I do} predicted
that the Yule tree prior will be selected as the best candidate model
more often in this scenario, which shows the sensitity of the method we used.

@@

We chose to use few taxa (as we preferred more replicates), but we do not
know the results would the data be more informative (that is, have more taxa).
A cool feature of \verb;pirouette; is that it can quantify when this happen;
when the prior is uninformative due to strong data.

>>>>

We \richel{(at least I. If you do not expect this, let me know the 
reasoning why)} \giovanni{I expect the same thing.} expected to find a bigger inference error 
when increasing the amount of co-occurring speciation events.
The reason is simple: the tree prior used assumes speciation does not
co-occur. The stronger the amount of co-occurring speciation, the stronger
the deviation from the species tree prior.

\giovanni{Further conclusions are to be drawn after results are produced.}

\iffalse % A model can always be made more accurate. There are two principles to follow when building one: accuracy, on one hand, is surely important but, at the same time, simplicity is important too as overcomplicated models miss the point of a research question. This is, for example, summarized in the request of a low AIC, which weights positively the accuracy of the model (high likelihood) and negatively its complexity (for which the number of parameters is a proxy).
\subsection{Research}

@@

\subsection{Methodological choices}
\fi

\iffalse % I choose to remove this too, at least for now, as it's too soon to write any conclusion when no result is available.
When simulating a DNA alignment, we chose to use the simplest (JC69) site
model and the simplest (strict) clock model. One could argue these models
are overly simplistic and biologically irrelevant. We argue that this
actually improves the clarity of our results for two reasons. The first
reason is a practical one: because the computational load was lowered,
we were able to perform more replicates. For a process with this high
amount of stochasticity, we think this is highly favored. The second reason
is that we think this reduces the noise of our results, without 
hurting the experiment: in our setup, it is essential
to assume the same generative site and clock model as the one that is actually 
used. Sure, we could have generated the alignment with a 
complex (GTR) site model and a complex (relaxed log-normal) clock model, then
correctly assume these were the generative site and clock model.
But that would be besides the point of the experiment: to measure the impact
of the tree prior.

When simulating our phylogenies, we used an MBD model with and without
extinction. In the case in which extinction is present,
picking the BD tree prior as the generative tree prior seems justified:
it is the best-fitting standard tree prior. For the MBD model without
extinction, \richel{@Giappo: I think this is an interesting insight:} 
we could have picked the Yule tree prior. We \richel{at least I do} predicted
that the Yule tree prior will be selected as the best candidate model
more often in this scenario, which shows the sensitivity of the method we used.

@@

We chose to use few taxa (as we preferred more replicates), but we do not
know the results would the data be more informative (that is, have more taxa).
A cool feature of \verb;pirouette; is that it can quantify when this happen;
when the prior is uninformative due to strong data.
\fi

### 2020-02-06 razzo-introduction.tex
Changes Giappo

We explore such case introducing the multiple-birth death model (MBD), 
currently absent in BEAST2. 
The multiple birth hypothesis aims to describe how 
large-scale environmental changes can lead phylogenies 
with a high abundance of speciation events in relatively short times.

>>>>

We explore such case introducing the multiple-birth death model (MBD), 
currently absent in BEAST2. 
The multiple birth (MB) hypothesis aims to describe how large-scale environmental changes can lead phylogenies with a high abundance of speciation events in relatively short times.

### 2020-02-06 razzo-methods.tex
Complete supplementary materials
Changes Giappo

We proceed in the following way:
under a range a parameter setting, 
we build simulated datasets generated under the multiple birth model. 
Then we run a \verb;pirouette; analysis, 
which will lead to error distributions
between the inference posterior and the original simulated trees. 
Importantly, the \verb;pirouette; analysis includes 
also a 'twin' parallel pipeline, 
which will provide a measure of the baseline error 
due to pure stochasticity, unavoidably occurring in the process.
We then analyse the complete and baseline errors.

\subsection{Model}

In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, 
to the common per-species speciation and extinction rates 
already present in the standard BD model. 
Additionally, MBD relies on two additional parameters, $\nu$ and $q$. 
The first, $\nu$, is the rate at which an environmental change is triggered.

When such event is triggered, each species present in the phylogeny 
at that moment has a probability $q$ to speciate at that time.
This kind of speciation is of a different nature with respect to 
the one triggered by $\lambda$. 
In fact, whereas parameter $\lambda$ can be seen as 
describing a sympatric process, $\nu$ induces the rise of 
geographical barriers that interrupt the gene flow,
resulting in allopatric speciation. 
Even though multiple speciations can co-occur, 
polytomies are not allowed in such process as each species can speciate only 
once during a co-occuring speciation event.

@@

@@

We ran multiple pilot experiments with 1 replicate to arrive at our final
parameter settings. We devised a set of rules to make a verdict about the
settings \richel{(script at \url{https://github.com/richelbilderbeek/razzo_project/blob/master/scripts/90_collect_run_times.R},
resulting verdicts at \url{https://github.com/richelbilderbeek/razzo_project/blob/master/verdict.md})}:
\itemize{
  \item quality: 95\% of all individual runs should have an ESS of at least 200.
  \item feasibility: 95\% of all individual runs should finish within 10 days.
  \item reproducibility: the mean run-time of all finished runs should be less than 24 hours 
    \richel{I suggest}.
  \item relevance: the percentage of taxa created by the MB process should be
    as high as possible
}

We searched through parameter space until these criteria were met.
All the parameter settings used in the pilot experiments can be found at 
\url{https://github.com/richelbilderbeek/razzo_project/blob/master/overview.md}.
Due to the low number of replicates, we were unable (nor tried)
to draw conclusions based on the results.

\subsection{Tree simulations}

We simulate the speciation process in continuous time 
using the \verb;mbd_sim; function from the \verb;mbd; R package (\citep{mbd}).
We let parameters vary using all possible combinations of values 
as shown in Table~\ref{tab:simulation_parameters}.
For each parameter setting, 
we generate [as much as time permits] 
\richel{Will need to calculate how many replicates we can do} 
independent reconstructed trees of the same crown age.

\subsection{Inference error estimation}

From each simulated 'true' MBD tree, we measure the impact of
ignoring the more complex and non-standard MBD tree prior in
Bayesian phylogentic inference.
We do so by assuming a simpler and standard BD tree prior,
and compare the 'true' tree with the inferred trees.
To do so, we use the \verb;pirouette; R package \citep{pirouette}.

In brief: \verb;pirouette; starts from 
a 'true' starting phylogeny (in our case: the simulated MBD tree), 
from which a DNA sequence alignment is simulated. 
From each sequence alignment, a Bayesian 
inference is run (in our case: with assuming a BD tree prior), 
to obtain a posterior distribution of jointly-estimated trees and model parameter estimates.
By comparing the true tree and the posterior trees, 
an inference error distribution is generated.
We use the twinning option available in \verb;pirouette;
that allows to quantify the impact of assuming a wrong
tree prior in a Bayesian inference.

In our experiment, the alignments are 1000 nucleotides in length, 
with a known root sequence of four 250 mono-nucleotide blocks, 
generated using the simplest nucleotide substitution model (JC69) 
and clock model (strict), with a mutation rate of $\frac{1}{2}\cdot c$,
in which $c$ is the crown age. 
With this mutation rate, each nucleotide has a 50\% chance
to mutate (both silently and non-silently) from the ancestral root sequence 
to any of the contemporary species' sequences at the tips.

For the Bayesian inference, we assume a generative model of 
a JC69 site model, a strict clock model and a BD tree prior. 
Additionally, we use an MRCA prior equal to the crown age with a normal distribution 
of width $\sigma = 0.01$. We pick an MCMC setup of $10^6$ states
with a sampling interval of once per $10^3$ states. Of the resulting
$10^3$ states, we discard a burn-in of $10\%$.

To verify that the generative model (with a BD tree prior) is
indeed the most appropriate standard model, we compare it to
39 other candidate inference models. These candidate models 
are all other combinations of four tree priors, 
two clock models and five speciation models, 
excluding the one combination used as a generative model.
We let the best candidate model also generate an error distribution,
which is expected to have a similar or high median.
For models comparison, \verb;pirouette; uses a nested sampling approach
to estimate marginal likelihoods,
as described in [\cite{maturana2018model}], 
using the \verb;NS; BEAST2 package. We setup the MCMC for the nested sampling
using its default values, which are a relative error $\epsilon$ of $10^-12$, 
1 particle, a sub-chain sampling interval of $5.10^3$ states with
a maximum chain length of $10^6$ states.

To compare the true tree with any of the posterior trees,
we use the absolute nLTT statistic \citet{janzen2015},
which will result in an error distribution with values
ranging from zero (when the inferred tree is identical 
to the true tree) to a maximum of one (trees are completely different).

@@

>>>>

We proceed in the following way: under a range of parameter setting, we build simulated datasets generated under the multiple birth model. 
Then we run a \verb;pirouette; analysis, which will lead to error distributions between the inferred posterior and the original simulated trees, for each setting. 
Importantly, the \verb;pirouette; analysis includes also a 'twin' parallel pipeline, which will provide a measure of the baseline error due to pure stochasticity, unavoidably occurring in the process.
We then analyse the complete and baseline errors.
\subsection{The Framework}

\subsubsection{Simulation Model}

In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, 
to the traditional per-species speciation and extinction rates 
already present in the standard BD model. 
Additionally, MBD relies on two additional parameters, $\nu$ and $q$. 
The first, $\nu$, is the rate at which an environmental change is triggered.

When such event is triggered, each species present in the phylogeny 
at that moment has a probability $q$ to speciate at that time.
This kind of speciation is of a different nature with respect to 
the one triggered by $\lambda$. 
In fact, whereas parameter $\lambda$ can be seen as 
describing a sympatric process, $\nu$ induces instead the rise of 
geographical barriers able to interrupt the gene flow in the population,
resulting in a possible allopatric speciation for each of the species.  
Even though multiple speciations can co-occur, 
polytomies are not allowed in such process as each species can speciate only 
once during a co-occuring speciation event.

@@

\subsubsection{Estimating the inference error}

From each simulated 'true' MBD tree, we measure the impact of
ignoring the more complex and non-standard MBD tree prior in
Bayesian phylogenetic inference.
We do so by assuming a simpler and standard BD tree prior,
and compare the 'true' tree with the inferred trees.
To do so, we use the \verb;pirouette; R package [\citep{pirouette}].

Here we summarize briefly how \verb;pirouette; works: it starts from a 'true' starting phylogeny (in our case: the simulated MBD tree), from which a DNA sequence alignment is simulated. 
From each sequence alignment, a Bayesian inference is run (in our case: with assuming a BD tree prior), to obtain a posterior distribution of jointly-estimated trees and model parameter estimates.
By comparing the true tree and the posterior trees, an inference error distribution is generated.
We use the twinning option available in \verb;pirouette; that allows to quantify the impact of assuming a wrong tree prior in a Bayesian inference compared to a reference background error.

@@

\iffalse
We ran multiple pilot experiments with 1 replicate to arrive at our final
parameter settings. We devised a set of rules to make a verdict about the
settings \richel{(script at \url{https://github.com/richelbilderbeek/razzo_project/blob/master/scripts/90_collect_run_times.R},
resulting verdicts at \url{https://github.com/richelbilderbeek/razzo_project/blob/master/verdict.md})}:
\itemize{
  \item quality: 95\% of all individual runs should have an ESS of at least 200.
    \richel{figure \ref{fig:esses}}
  \item feasibility: 95\% of all individual runs should finish within 10 days.
    \richel{figure \ref{fig:runtimes}}
  \item reproducibility: the mean run-time of all finished runs should be less than 24 hours 
    \richel{I suggest}.
    \richel{figure \ref{fig:runtimes}}
  \item relevance 1: the percentage of taxa created by the MB process should be
    as high as possible
  \item relevance 2: the percentage of taxa should be
    as high as possible
}
We searched through parameter space until these criteria were met.
All the parameter settings used in the pilot experiments can be found at 
\url{https://github.com/richelbilderbeek/razzo_project/blob/master/overview.md}.
Due to the low number of replicates, we were unable (nor tried)
to draw conclusions based on the results.
\fi

When working on simulated datasets the choice of the generating parameters is always an arbitrary choice. In principle one would choose a parameter range as wide as possible to characterize the most of the available parameter space. However, there are technical aspects that must be taken in consideration to be sure that both the simulation and the inference processes work out in an optimal way.
We required the following criteria to be satisfied \giovanni{Does this apply only to the mbd parameters or also to the pirouette ones?}:
\itemize{
  \item Quality: At least 95\% of all individual runs should have an Estimated Sample Size (ESS) of at least 200 (see \cite{beast} for reference);
    \richel{figure \ref{fig:esses}}
  \item Feasibility: As time is a limited resource we impose that at least 95\% of all individual runs should finish within 10 days;
    \richel{figure \ref{fig:runtimes}}
  \item Reproducibility: the mean run-time of all finished runs should be less than 24 hours \richel{figure \ref{fig:runtimes}}

    \richel{I suggest}. \giovanni{What's the difference with the previous point?}
  \item Relevance 1: In order to have a meaningful inference, we demand that the MB signal is detectable to a satisfactory extent. For this reason the percentage of taxa created by the MB process should be as high as possible;
  \item Relevance 2: Small trees might not contain a significant amount of information to be analyzed. For this reason we demand that the number of taxa should be as high as possible, as long as all the other requirements are met;
}

Additional information about the choice of a suitable parameter setting can be found in the Supplementary Material.
\iffalse
\giovanni{This could be put in the appendix, if we manage to put in a handy format.}
We searched through parameter space until these criteria were met.
All the parameter settings used in the pilot experiments can be found at 
\url{https://github.com/richelbilderbeek/razzo_project/blob/master/overview.md}.
Due to the low number of replicates, we were unable (nor tried)
to draw conclusions based on the results.
\fi

\subsubsection{Simulations setting}

We simulate the speciation process in continuous time using the \verb;mbd_sim; function from the \verb;mbd; R package [\citep{mbd}].
We let parameters vary using all possible combinations of values as shown in Table~\ref{tab:simulation_parameters}.
For each parameter setting, we generate [as much as time permits] 
\richel{Will need to calculate how many replicates we can do} 
independent trees of the same crown age. Note that trees that are actually passed as \verb;pirouette; input are not full trees, but reconstructed trees instead. A reconstructed tree is a tree where all the species that went extinct before the present time are removed, as this information is usually unavailable on real data phylogenies.

\input{tab_simparameters.tex}

\subsubsection{pirouette setting}

In our experiment, the alignments are 1000 nucleotides in length, with a known root sequence of four 250 mono-nucleotide blocks, generated using the simplest nucleotide substitution model (Jukes Cantor, JC69) and clock model (strict), with a mutation rate of $\frac{t_c}{2}$, where $t_c$ is the crown age. 
With this mutation rate, each nucleotide has a 50\% chance to mutate (both silently and non-silently) from the ancestral root sequence to any of the contemporary species' sequences at the tips.

For the Bayesian inference, we assume a generative model of a JC69 site model, a strict clock model and a BD tree prior.
Additionally, we use a Most Recent Common Ancestor (MRCA) prior equal to the crown age with a normal distribution of width $\sigma = 0.01$. We pick a Markov Chain Monte Carlo (MCMC) setup of $10^6$ states with a sampling interval of once per $10^3$ states. Of the resulting $10^3$ states, we discard a burn-in of $10\%$.

To verify that the generative model (with a BD tree prior) is indeed the most appropriate standard model, we compare it to 39 other candidate inference models.
These candidate models are all other combinations of four tree priors, two clock models and five speciation models, excluding the one combination used as a generative model.
We let the best candidate model also generate an error distribution, which is expected to have a similar or high median.
For models comparison, \verb;pirouette; uses a nested sampling approach to estimate marginal likelihoods, as described in \citet{russel2019model}, using the "\verb;NS;" BEAST2 package. We setup the MCMC for the nested sampling using its default values, which are a relative error $\epsilon$ of $10^{-12}$, 
1 particle, a sub-chain sampling interval of $5 \cdot 10^3$ states with a maximum chain length of $10^6$ states.

To compare the true tree with any of the posterior trees, we use the absolute nLTT statistic by \citet{janzen2015}, which will result in an error distribution with values ranging from zero (when the inferred tree is identical to the true tree) to a maximum of one (trees are completely different).

@@

\giovanni{Add mention to Jensen-Shannon divergence metric, if used.}
